---
title: "Package development"
output: html_document
---

## Introduction
This file contains code and annotations that are useful when developing a package.

## Install packages used in this file
```{r prepare, eval = FALSE}
required_pkgs <- c("devtools", "tinytest", "usethis")
bool_install <- !vapply(X = required_pkgs, FUN = requireNamespace,
                        FUN.VALUE = logical(1), quietly = TRUE)
pkgs_install <- required_pkgs[bool_install]
if(length(pkgs_install) > 0L) {
  utils::install.packages(pkgs = pkgs_install, lib = NULL, dependencies = NA,
                          quiet = FALSE)
}
```

## Add functions
### Set up
```{r functions, eval = FALSE}
usethis::use_r("is_number")
devtools::document() # also runs devtools::load_all()
?is_number() # view help-page of the function
```

Do:

- Manually update the `NEWS`-file.

### Documentation
```{r documentation, eval = FALSE}
#' Title
#'
#' Description
#'
#' @inheritParams is_logical # From a function in the current package
#' @inheritParams utils::installed.packages # from a function in another package
#' @param x Vector of names to test.
#' @param allow_underscores `TRUE` or `FALSE`: allow underscores?
#' @param x,y Separate arguments by a comma without a space to create a single
#' description for multiple arguments.
#'
#' @details
#'
#' @returns
#' `TRUE` or `FALSE`, returned [invisibly][invisible].
#'
#' @section Side effects:
#' Values in `x` are identified in an [error][stop], [warning], or [message] if
#' `signal` is `error`, `warn`, or `message`, respectively. Set `signal` to
#' `quiet` to suppress this.
#'
#' @section Notes:
#' Some notes.
#'
#' @section Programming notes:
#' Some programming notes. Create a GitHub issue instead of To do / Wishlist
#' sections.
#'
#' @seealso
#' Section `Details` of [make.names()] and the [\R FAQ about valid names](
#' https://CRAN.R-project.org/doc/manuals/R-FAQ.html#What-are-valid-names_003f)
#' on the syntactical validity of names.
#'
#' @family collections of checks on type and length
#'
#' @examples
#' all_names(x = c("a", "b2a")) # TRUE
#'
#' @export
func_name <- function(x, allow_underscores = TRUE) {
  stopifnot(checkinput::is_logical(allow_underscores))
  ...
}
```


#### Inherit sections
Inherited sections are *silently* ignored if that section is also defined in the
file itself.

``` {r inherit docs, eval = FALSE}
#' @inherit is_number return
#' @inherit is_number details
#' @inheritSection is_logical Programming notes
#' @inheritSection is_logical @note
```

### Add tests
``` {r tests, eval = FALSE}
library(tinytest)
tinytest::test_all() # Run all tests of a package
tinytest::run_test_file("./inst/tinytest/test_funcname.R") # Run specific file

# expect_silent(...) tests that no warnings or errors are emitted, NOT that no
# messages are emitted.
expect_silent(expect_true(func_name(x = x, arg = arg)))
expect_warning(
  expect_equal(func_name(x = "a", arg = arg), 3),
  pattern = "...", strict = TRUE, fixed = TRUE)
expect_error(func_name(x = "a", arg = arg),
             pattern = "is_number(x) is not TRUE", fixed = TRUE)
```

### Updating dependencies
``` {r dependencies, eval = FALSE}
# Declare a minimum version for R itself
usethis::use_package("R", type = "Depends", min_version = "4.1.0")

# Declare a dependency on a package: 'use_package()' to declare a minimum
# version and 'use_dev_package()' to specify the remote repository to download
# the package from.
usethis::use_package(package = "checkinput", type = "Imports", min_version = TRUE)
usethis::use_dev_package(package = "checkinput", type = "Imports",
                         remote = "github::JesseAlderliesten/checkinput")
```

Do:

- Specify the minimum declared `R` version as workflow in GitHub Actions, see
  the section `Use GitHub Actions`.

To use a function that is present in 'new' versions of R also in older versions
of R, see https://github.com/yihui/xfun/blob/main/R/zzz.R

#### Importing packages or functions
There are two ways to use functions from packages that are imported (i.e., are
in the `Imports:` field of the `DESCRIPTION` file): (1) use the package name
followed by two colons and the function name, e.g., `utils::osVersion(...)`;
(2) In addition to the `Imports:` field of the `DESCRIPTION` file, list the
function in the `NAMESPACE` file, e.g., `importFrom(utils, osVersion)`. With
`usethis`, this can be achieved by the following line:
`usethis::use_import_from(package = "utils", fun = "osVersion")`. That adds the
line `#' @importFrom utils osVersion` in `R/<pkgname>-package.R` that was
created by `usethis::use_package_doc()`, and adds the line `Imports: utils` to
the `DESCRIPTION` file (you might have to use `devtools::document()` for this to
take effect).

``` {r imports, eval = FALSE}
usethis::use_import_from(package = "BiocManager", fun = "valid")
usethis::use_package("tinytest", type = "Suggests", min_version = "1.4.1")
```

## Add vignettes
### Setting up
``` {r vignettes, eval = FALSE}
usethis::use_vignette("my_vignette", title = "Some title")
usethis::use_package(package = "knitr", type = "Suggests")
usethis::use_package(package = "rmarkdown", type = "Suggests")
devtools::document()
devtools::build_vignettes()
browseVignettes(package = basename(getwd()))
```

If no vignettes are visible, `devtools::install()` was probably run with the
default argument `build_vignettes = FALSE`, use `build_vignettes = TRUE` instead:
`devtools::install(quick = FALSE, upgrade = "never", build_vignettes = TRUE)`.

See also `tools::pkgVignettes(package = "checkrpkgs")` and
`utils::vignette(<"vignette_title">)`. To render a `Rd`-file as plain-text help
in the R console: `tools::Rd2txt("path/to/file.Rd")`.

### Styling
Put the next lines in the header of vignettes to get a table of contents:
``` {r vignette style, eval = FALSE}
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
```
See also my Chrome Bookmarks: `R` > `CreatePkgs` > `RMarkdown`.

### Linking
For internal links to other sections in the same document ('Section', 'above',
'below'): `[<Section title>]` or `[<link text>][<Section title>]`

To link from vignettes to help-pages in the same package, use
`[someFun](../help/someFun)` (from a
[StackOverflow answer](https://stackoverflow.com/questions/34946219)).

Text between angled brackets (e.g., `<pkgname>`) disappears when used in link
text, so instead of `[library(<pkgname>)](../help/library)`, use

``` {r link, eval = FALSE}
[library](../help/library)`(<pkgname>)`
```

For a more advanced way of linking from vignettes to help-pages see
[here](https://github.com/dmurdoch/rgl/commit/bbc84447c2a6efed42907fbac176e9569b868d8f).

To link from help-pages to vignettes, use:
The `[vignette about <some description>](../doc/<filename>.html)`.
If no vignettes are visible, run `devtools::build_vignettes()`, which possibly
needs to be followed by `install(..., build_vignettes = TRUE)`.


## Update package-wide documentation

### Updating the README
To provide a nice overview of functions: see
[this example](https://github.com/MJobinASU/MontyHall). After adjusting the
`README.Rmd`, `Knit` it to produce a `README.Md` file.


## Use GitHub Actions
To manually run GitHub Actions (GHA) set up `workflow_dispatch` (see section
`Use GitHub Actions` in `pkg_setup.Rmd`). To check if package `B`, which depends on
package `A`, is still working fine, go to the `Actions` tab of repository `B`,
select the action you want to trigger (e.g., `R-CMD-check.yaml`), and use the
`Run workflow` button to run the GHA. You can select which branch it should
run on, but you need to trigger it once manually on the `main` branch to be
able to trigger it manually on other branches.

Scheduled jobs that failed can be rerun through the button `Re-run jobs` >
`Re-run failed jobs` > `Re-run jobs`.

If the package declares a dependency on a specific `R` version, it is useful to
specify the minimum declared `R` version to run in addition to the ones that are
by default used in the template: add `- {os: ubuntu-latest,   r: '4.0.0'}` to
the section `matrix: config:` to run `R` 4.0.0.

## Using pkgdown
See https://r-pkgs.org/website.html and https://pkgdown.r-lib.org/

Updating website
``` {r pkgdown, eval = FALSE}
pkgdown::build_site()
```
Open `docs/index.html` in a webbrowser to preview the website, or look at the
files that constitute your package's website are in the local `docs/` directory.
