---
title: "Package development"
output: html_document
---

## Introduction
This file contains code and annotations that are useful when developing a package.

## Install packages used in this file
```{r prepare, eval = FALSE}
required_pkgs <- c("devtools", "tinytest", "usethis")
bool_install <- !vapply(X = required_pkgs, FUN = requireNamespace,
                        FUN.VALUE = logical(1), quietly = TRUE)
pkgs_install <- required_pkgs[bool_install]
if(length(pkgs_install) > 0L) {
  utils::install.packages(pkgs = pkgs_install, lib = NULL, dependencies = NA,
                          quiet = FALSE)
}
```

## Add functions
### Set up
```{r functions, eval = FALSE}
usethis::use_r("is_number")
devtools::document() # also runs devtools::load_all()
?is_number() # view help-page of the function
```

Do:

- Manually update the `NEWS`-file.

### Documentation
```{r documentation, eval = FALSE}
#' Title
#'
#' Description
#'
#' @inheritParams is_logical # From a function in the current package
#' @inheritParams utils::installed.packages # from a function in another package
#' @param x Vector of names to test.
#' @param allow_underscores `TRUE` or `FALSE`: allow underscores?
#' @param x,y Separate arguments by a comma without a space to create a single
#' description for multiple arguments.
#'
#' @details
#'
#' @returns
#' `TRUE` or `FALSE`, returned [invisibly][invisible].
#'
#' @section Side effects:
#' Values in `x` are identified in an [error][stop], [warning], or [message] if
#' `signal` is `error`, `warn`, or `message`, respectively. Set `signal` to
#' `quiet` to suppress this.
#'
#' @section Notes:
#' Some notes.
#'
#' @section Programming notes:
#' Some programming notes. Create a GitHub issue instead of To do / Wishlist
#' sections.
#'
#' @seealso
#' Section `Details` of [make.names()] and the [\R FAQ about valid names](
#' https://CRAN.R-project.org/doc/manuals/R-FAQ.html#What-are-valid-names_003f)
#' on the syntactical validity of names.
#'
#' @family collections of checks on type and length
#'
#' @examples
#' all_names(x = c("a", "b2a")) # TRUE
#'
#' @export
func_name <- function(x, allow_underscores = TRUE) {
  stopifnot(checkinput::is_logical(allow_underscores))
  ...
}
```


#### Inherit sections
Inherited sections are *silently* ignored if that section is also defined in the
file itself.

``` {r inherit docs, eval = FALSE}
#' @inherit is_number return
#' @inherit is_number details
#' @inheritSection is_logical Programming notes
#' @inheritSection is_logical @note
```

### Add tests
``` {r tests, eval = FALSE}
library(tinytest)
tinytest::test_all() # Run all tests of a package
tinytest::run_test_file("./inst/tinytest/test_funcname.R") # Run specific file

# expect_silent(...) tests that no warnings or errors are emitted, NOT that no
# messages are emitted.
expect_silent(expect_true(func_name(x = x, arg = arg)))
expect_warning(
  expect_equal(func_name(x = "a", arg = arg), 3),
  pattern = "...", strict = TRUE, fixed = TRUE)
expect_error(func_name(x = "a", arg = arg),
             pattern = "is_number(x) is not TRUE", fixed = TRUE)
```

### Using temporary files
- See `testthat::test_path()` on changing paths depending on how a check is run.
- On using temporary files, see `local(...)`,
  [this](https://stat.ethz.ch/pipermail/r-devel/2018-March/075783.html) R-mail,
  and `withr::with_tempfile()`:
```{r tempfile, eval = FALSE}
\dontshow{
  my_tempdir <- tempdir()
  example_dir <- file.path(my_tempdir, example_dir)
}
```

### Update dependencies
``` {r dependencies, eval = FALSE}
# Declare a minimum version for R itself
usethis::use_package("R", type = "Depends", min_version = "4.1.0")

# Declare a dependency on a package: 'use_package()' to declare a minimum
# version and 'use_dev_package()' to specify the remote repository to download
# the package from.
usethis::use_package(package = "checkinput", type = "Imports", min_version = TRUE)
usethis::use_dev_package(package = "checkinput", type = "Imports",
                         remote = "github::JesseAlderliesten/checkinput")
```

Do:

- Specify the minimum declared `R` version as workflow in GitHub Actions, see
the section `Use GitHub Actions`.

To use a function that is present in 'new' versions of R also in older versions
of R, see https://github.com/yihui/xfun/blob/main/R/zzz.R

#### Importing packages or functions
The standard way to use functions from other packages is to import the package
(i.e., put it in the `Imports:` field of the `DESCRIPTION` file) and in code use
the package name followed by two colons and the function name, e.g.,
`utils::osVersion(...)`. If that is not possible, for example because the
function is an operator, it is necessary to also list the function in the
`NAMESPACE` file, e.g., to add the line `#' @importFrom utils osVersion` to
`R/<pkgname>-package.R` that was created by `usethis::use_package_doc()`. This
is done by `usethis::use_import_from(package = "utils", fun = "osVersion")` (you
might have to use `devtools::document()` for this to take effect).

``` {r imports, eval = FALSE}
usethis::use_import_from(package = "BiocManager", fun = "valid")
usethis::use_package("tinytest", type = "Suggests", min_version = "1.4.1")
```

## Add vignettes
### Set up
``` {r vignettes, eval = FALSE}
usethis::use_vignette("my_vignette", title = "Some title")
usethis::use_package(package = "knitr", type = "Suggests")
usethis::use_package(package = "rmarkdown", type = "Suggests")
devtools::document()
devtools::build_vignettes()
browseVignettes(package = basename(getwd()))
```

If no vignettes are visible, `devtools::install()` was probably run with the
default argument `build_vignettes = FALSE`, use `build_vignettes = TRUE` instead:
`devtools::install(quick = FALSE, upgrade = "never", build_vignettes = TRUE)`.

See also `tools::pkgVignettes(package = "checkrpkgs")` and
`utils::vignette(<"vignette_title">)`. To render a `Rd`-file as plain-text help
in the R console: `tools::Rd2txt("path/to/file.Rd")`.

### Styling
Put the next lines in the header of vignettes to get a table of contents:
``` {r vignette style, eval = FALSE}
output:
  rmarkdown::html_vignette:
  toc: true
toc_depth: 3
```
See also my Chrome Bookmarks: `R` > `CreatePkgs` > `RMarkdown`.

### Linking
For internal links to other sections in the same document ('Section', 'above',
'below'): `[<Section title>]` or `[<link text>][<Section title>]`

To link from vignettes to help-pages in the same package, use
`[someFun](../help/someFun)` (from a
[StackOverflow answer](https://stackoverflow.com/questions/34946219)).

Text between angled brackets (e.g., `<pkgname>`) disappears when used in link
text, so instead of `[library(<pkgname>)](../help/library)`, use

``` {r link, eval = FALSE}
[library](../help/library)`(<pkgname>)`
```

For a more advanced way of linking from vignettes to help-pages see
[here](https://github.com/dmurdoch/rgl/commit/bbc84447c2a6efed42907fbac176e9569b868d8f).

To link from help-pages to vignettes, use:
The `[vignette about <some description>](../doc/<filename>.html)`.
If no vignettes are visible, run `devtools::build_vignettes()`, which possibly
needs to be followed by `install(..., build_vignettes = TRUE)`.

## Preparing for updates

### Check tests
Check for which functions no test file has been written.

``` {r update tests, eval = FALSE}
devtools::document()
tinytest::test_all()

files_R <- list.files(file.path(".", "R"))
files_man <- list.files(path = file.path(".", "man"), pattern = "\\.Rd$")
expected_man <- sub(pattern = "\\.Rd$", replacement = ".R", x = files_man)
expected_files <- sort(unique(paste0("test_", c(files_R, expected_man))))
expected_files[!(expected_files %in% list.files("./inst/tinytest"))]
```

### Locally test update
``` {r update, eval = FALSE}
devtools::document()
# Using `manual = FALSE` because building the manual leads to `latex` errors
devtools::check(manual = FALSE, force_suggests = TRUE, run_dont_test = TRUE)
.libPaths() # Check if output of .libPaths() is correct.
devtools::install(quick = FALSE, upgrade = "ask", build_vignettes = TRUE)
devtools::install(quick = FALSE, upgrade = "never", build_vignettes = TRUE)

# Load the package and view the help files as usual outside devtools:
library(basename(getwd()), character.only = TRUE)
browseVignettes(package = basename(getwd()))
?reorder_levels
```

If no vignettes are visible, `devtools::install()` was probably run with the
default `build_vignettes = FALSE`: change it to use `build_vignettes = TRUE`.

### Update package-wide documentation

#### NEWS
Restyle and publish the `NEWS`-file at each new package release.

#### README
To provide a nice overview of functions: see
[this example](https://github.com/MJobinASU/MontyHall). After adjusting the
`README.Rmd`, `Knit` it to produce a `README.Md` file.

#### Increment package version
To increment the package version, adjust the package version in the `DESCRIPTION`
file, or run `R` as administrator and then use `usethis::use_version()`. In the
latter case, do not automatically `commit` the change, but do so manually to
adjust the commit message to something like
`Bump to version x.y.z. Breaking change: <func> no longer ...`. Indicating the
version number in the commit title makes it easier to find back changes later on.

### Use GitHub Actions
To manually run GitHub Actions (GHA) set up `workflow_dispatch` (see section
`Use GitHub Actions` in `pkg_setup.Rmd`). To check if package `B`, which depends
on package `A`, is still working fine, go to the `Actions` tab of repository `B`,
select the action you want to trigger (e.g., `R-CMD-check.yaml`), and use the
`Run workflow` button to run the GHA. You can select which branch it should
run on, but you need to trigger it once manually on the `main` branch to be
able to trigger it manually on other branches.

Scheduled jobs that failed can be rerun through the button `Re-run jobs` >
`Re-run failed jobs` > `Re-run jobs`.

If the package declares a dependency on a specific `R` version, it is useful to
specify the minimum declared `R` version to run in addition to the ones that are
by default used in the template: add `- {os: ubuntu-latest,   r: '4.0.0'}` to
the section `matrix: config:` to run `R` 4.0.0.

### Update pkgdown website
See https://r-pkgs.org/website.html and https://pkgdown.r-lib.org/

Updating website
``` {r pkgdown, eval = FALSE}
pkgdown::build_site()
```
Open `docs/index.html` in a webbrowser to preview the website, or look at the
files that constitute your package's website are in the local `docs/` directory.

### Merge devel-branch with master
To merge the `devel`-branch with `master`, go the the `devel`-branch on GitHub.
`<Contribute>` > `Open Pull Request`. Copy the updated `NEWS` in the
`description` field and use the button `Create pull request`.

If you see
`No conflicts with base branch: Merging can be performed automatically`, you can
use the green `Merge pull request` button, or change to `Squash and merge`.

Otherwise, you will see `This branch has conflicts that must be resolved`:
follow the instructions to resolve the conflicts. Then push the button
`Commit merge`. Then you should see `No conflicts with base branch` and you can
proceed as described in the previous paragraph.

After a successful merge, you will see a message that you can delete the
`devel`-branch, which you can do. To do so later, go to `Pull requests`, select
the `closed` tab, and scroll down to the button `Delete branch`.

In `RStudio`, open the relevant project to check there are no commits left. Then
you can move to the `master` branch and `Pull` to get all updates you just
committed to the `master` branch, or first delete the project and `Pull` the
fresh one.

Then you can delete the old package versions (see `getwd()` for its location)
and install the updated versions following your own instructions on the GitHub
pages of the relevant packages.

### Set up a new GitHub branch
Then set up a new GitHub branch:
- At the GitHub page of the package:
`Branch` icon > green `New branch` button > use `devel` as branch name >
green `Create new branch` button.
- Go back to the GitHub page of the package, and at the green `Code` button >
`copy URL to clipboard`.
- In `RStudio`: `File` > `New project` > `Version control` > `Git` and paste the
copied URL in the `Repository URL` > `Create Project`.
- Then (still in `RStudio`) in the `Git` menu change from `master` to the
`devel`-branch.

## Troubleshooting
To prevent `regex`-classes in example code from being interpreted as links
(which leads to the error
`@section Could not resolve link to topic ":blank:" in the dependencies or base packages`)
when running `devtools::document()`, use backticks (`) to format a line as code,
or wrap consecutive lines in `\code{...}`.

## Guidelines
See https://developer.r-project.org/ and 
https://developer.r-project.org/devel-guidelines.txt
