---
title: "Package development"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Package development}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction
This file contains code and annotations that are useful when developing a package.

## Add functions
### Set up
```{r functions, eval = FALSE}
usethis::use_r("is_number")
devtools::document() # also runs devtools::load_all()
?checkinput::is_number() # view help-page of the function
```

Do:

- Manually update the `NEWS`-file.

### Documentation
```{r documentation, eval = FALSE}
#' Title
#'
#' Description
#'
#' # Since `roxygen2` 8.0.0, it is possible to specify which arguments to inherit
#' @inheritParams is_logical allow_zero # From a function in the current package
#' @inheritParams utils::installed.packages # from a function in another package
#' @param x Vector of names to test.
#' @param allow_underscores `TRUE` or `FALSE`: allow underscores?
#' @param x,y Separate arguments by a comma without a space to create a single
#' description for multiple arguments.
#'
#' @details
#'
#' @returns
#' `TRUE` or `FALSE`, returned [invisibly][invisible].
#'
#' @section Side effects:
#' Values in `x` are identified in an [error][stop], [warning], or [message] if
#' `signal` is `error`, `warn`, or `message`, respectively. Set `signal` to
#' `quiet` to suppress this.
#'
#' @section Notes:
#' Some notes.
#'
#' @section Programming notes:
#' Some programming notes. Create a GitHub issue instead of To do / Wishlist
#' sections.
#'
#' @seealso
#' Section `Details` of [make.names()] and the [\R FAQ about valid names](
#' https://CRAN.R-project.org/doc/manuals/R-FAQ.html#What-are-valid-names_003f)
#' on the syntactical validity of names.
#'
#' @family collections of checks on type and length
#'
#' @examples
#' all_names(x = c("a", "b2a")) # TRUE
#'
#' @export
func_name <- function(x, allow_underscores = TRUE) {
  stopifnot(checkinput::is_logical(allow_underscores))
  ...
}
```


#### Inherit sections
Inherited sections are *silently* ignored if that section is also defined in the
file itself.

``` {r inherit docs, eval = FALSE}
#' @inherit is_number return
#' @inherit is_number details
#' @inheritSection is_logical Programming notes
#' @inheritSection is_logical @note
```

#### Add examples
Examples should **not** write to the working directory (i.e., use `getwd()`) but
to a temporary directory that is cleaned up afterwards. See the section
`Usage in practice` in `help("create_tempdir", package = "progutils")`.

Examples rendered on a website created with [pkgdown](https://pkgdown.r-lib.org/)
will display the output underneath the code. However, help-files accessed
through `help(<func>)` will not, such that it might be useful to include a short
comment regarding what feature of the output is notable.

### Add tests
Tests should **not** write to the working directory (i.e., use `getwd()`) but to
a temporary directory that is cleaned up afterwards. See the section
`Usage in practice` in `help("create_tempdir", package = "progutils")`.

``` {r tests, eval = FALSE}
library(tinytest)
tinytest::test_all() # Run all tests of a package
tinytest::run_test_file("./inst/tinytest/test_funcname.R") # Run specific file

# expect_silent(...) tests that no warnings or errors are emitted, NOT that no
# messages are emitted.
expect_silent(expect_true(func_name(x = x, arg = arg)))
expect_warning(
  expect_equal(func_name(x = "a", arg = arg), 3),
  pattern = "...", strict = TRUE, fixed = TRUE)
expect_error(func_name(x = "a", arg = arg),
             pattern = "is_number(x) is not TRUE", fixed = TRUE)
```

### Update dependencies
Specifying a minimum package version when importing packages ensures that used
features are actually present. Setting `min_version = TRUE` in 
`usethis::use_package()` uses the currently installed package version.
Alternatively, check which features are used and in which version those were
introduced to find a minimum version. That can also be done by specifying
minimum versions in checks through GitHub actions.

``` {r dependencies, eval = FALSE}
# Declare a minimum version for R itself
usethis::use_package("R", type = "Depends", min_version = "4.1.0")

# Declare a dependency on a package: 'use_package()' to declare a minimum
# version and 'use_dev_package()' to specify the remote repository to download
# the package from.
usethis::use_package(package = "checkinput", type = "Imports", min_version = TRUE)
usethis::use_dev_package(package = "checkinput", type = "Imports",
                         remote = "github::JesseAlderliesten/checkinput")
```

Do:

- Specify the minimum declared `R` version as workflow in GitHub Actions, see
the section [Use GitHub Actions].

To use a function that is present in 'new' versions of R also in older versions
of R, see https://github.com/yihui/xfun/blob/main/R/zzz.R

#### Importing packages or functions
The standard way to use functions from other packages is to import the package
(i.e., put it in the `Imports:` field of the `DESCRIPTION` file) and in code use
the package name followed by two colons and the function name, e.g.,
`utils::osVersion()`. If that is not possible, for example because the function
is an operator, it is necessary to also list the function in the `NAMESPACE`
file, e.g., to add the line `#' @importFrom utils osVersion` to
`R/<pkg>-package.R` that was created by `usethis::use_package_doc()`. This
is done by `usethis::use_import_from(package = "utils", fun = "osVersion")`.

``` {r imports, eval = FALSE}
usethis::use_import_from(package = "utils", fun = "osVersion")
usethis::use_package("utils", type = "Suggests", min_version = "4.1.0")
```

## Add vignettes
### Set up
``` {r vignettes, eval = FALSE}
usethis::use_vignette("my_vignette", title = "Some title")
usethis::use_package(package = "knitr", type = "Suggests")
usethis::use_package(package = "rmarkdown", type = "Suggests")
devtools::document()
pkgdown::build_article()
browseVignettes(package = basename(getwd()))
```

If no vignettes are visible, `devtools::install()` was probably run with the
default argument `build_vignettes = FALSE`, use `build_vignettes = TRUE` instead:
`devtools::install(quick = FALSE, upgrade = FALSE, build_vignettes = TRUE)`. See
also `tools::pkgVignettes(package = "<pkg>")` for information about which
vignettes R finds.

Use `utils::vignette("<vignette_title>")` (e.g., `utils::vignette("pkg_devel")`)
and `tools::Rd2txt("path/to/file.Rd")` to render a vignette or an `Rd`-file in
the help-pane of RStudio, respectively.

### Styling
See the vignette *RMarkdown and knitr*:
`vignette("rmarkdown_knitr", package = "develcoder")`) on using RMarkdown to
style vignettes.

Put the next lines in the header of vignettes to get a table of contents:
``` {r vignette style, eval = FALSE}
output:
  rmarkdown::html_vignette:
  toc: true
toc_depth: 3
```

### Linking
For internal links to other sections in the same document ('Section', 'above',
'below'): `[<Section title>]` or `[<link text>][<Section title>]`.

Linking to a help page of a package: `help("<func>", package = "<pkg>")`
(`help("<func>::<pkg>")` does **not** work).

There is [**no**](https://r-pkgs.org/vignettes.html#links) official way to link
from vignettes to help-pages. Using relative paths (e.g.,
`[<func>](../help/<func>)`, as proposed in a
[StackOverflow answer](https://stackoverflow.com/questions/34946219)) is
impaired by the change in the location of files when the package is installed.
In addition, linking on the `pkgdown` website does **not** work in that format
(`pkgdown` recognises calls like `` `<func>()` `` and
`` `<pkg>::<func>()` ``, see the
[documentation](https://pkgdown.r-lib.org/articles/linking.html)). For a more
advanced way of linking from vignettes to help-pages that might work, see
[here](https://github.com/dmurdoch/rgl/commit/bbc84447c2a6efed42907fbac176e9569b868d8f).

To link from help-pages to vignettes, use
``` {r link vignettes, eval = FALSE}
The vignette *<vignette>*: `vignette("<vignette>", package = "<pkg>")` 
```

## Adding miscellaneous files
Non-standard files or folders should be added in the `inst/` directory to pass
`R cmd check`. Those files and folders will be in the top directory in the
installed package.

## Preparing for updates

### Check tests
Check for which functions no test file has been written.

``` {r update tests, eval = FALSE}
devtools::document()
tinytest::test_all()

curr_pkg_name <- basename(getwd())
files_R <- list.files(file.path(".", "R"))
files_man <- list.files(path = file.path(".", "man"), pattern = "\\.Rd$")
proj_pkg_R <- paste0(curr_pkg_name, "-package.R")
if(proj_pkg_R %in% files_R) {
  files_R <- files_R[files_R != proj_pkg_R]
  files_man <- files_man[files_man != paste0(proj_pkg_R, "d")]
}
if("reexports.Rd" %in% files_man) {
  files_man <- files_man[files_man != "reexports.Rd"]
  # Based on https://stackoverflow.com/a/74487073/32365738
  reexported_funcs <- progutils::not_in(getNamespaceExports(curr_pkg_name),
                                        ls(envir = asNamespace(curr_pkg_name)))
  if(length(reexported_funcs) == 0L) {
    warning("File 'reexports.Rd' is present but no re-exported functions found!")
  } else {
    files_R <- progutils::not_in(files_R, paste0(reexported_funcs, ".R"))
  }
}
expected_man <- sub(pattern = "\\.Rd$", replacement = ".R", x = files_man)
expected_files <- sort(unique(paste0("test_", c(files_R, expected_man))))
tests_missing <- expected_files[!(expected_files %in% list.files("./inst/tinytest"))]
if(length(tests_missing) > 0L) {
  warning("No test-file found for file ", progutils::paste_quoted(tests_missing))
}
```

### Automated checks
``` {r checks, eval = FALSE}
devtools::document()
# Using `manual = FALSE` because building the manual leads to `latex` errors
devtools::check(manual = FALSE, force_suggests = TRUE, run_dont_test = TRUE)
devtools::release_checks()
if(utils::packageVersion("devtools") >= "2.5.0") {
  # check for missing `\value` and `\examples` fields in `Rd` files
  devtools::check_doc_fields()
}

tools::check_package_urls(dir = getwd())
tools::check_package_dois(dir = getwd())

pkg_env <- devtools::load_all()$env
codetools::checkUsageEnv(env = pkg_env, all = TRUE, suppressParamAssigns = TRUE)

# - 'pkgcheck' depends on 'pkgstats', which needs the system libraries
#   'ctags-universal' and 'GNU global'. Run pkgstats::ctags_test() to check if
#   those are installed. If not, you might have to run pkgstats::ctags_install();
#   see also the output of library(pkgstats) and the documentation at
#   https://docs.ropensci.org/pkgstats/articles/installation.html
# - pkgcheck::pkgcheck() and goodpractice::gp() have quite some overlap. See
#   https://github.com/ropensci-review-tools/goodpractice/issues/284 
# - The lintr issue `Use == instead of %in% for scalar comparison` should be
#   ignored for 'x %in% y' if 'x' or 'y' might contain 'NA' that should be
#   removed: 'x %in% NA' and 'NA %in% y' return 'FALSE', whereas 'x == NA' and
#   'NA == y' return 'NA'.
res_pkgcheck <- pkgcheck::pkgcheck()
res_pkgcheck
res_goodpractice <- goodpractice::gp()
res_goodpractice

# Check spelling (requires system library like Aspell, see http://aspell.net/)
utils::aspell_package_Rd_files(dir = getwd())
utils::aspell_package_vignettes(dir = getwd())
```

### Locally test update
``` {r update, eval = FALSE}
devtools::document()
.libPaths() # Check if output of .libPaths() is correct.
devtools::install(quick = FALSE, upgrade = FALSE, build_vignettes = TRUE)

# Load the package and view the help files as usual outside devtools:
library(basename(getwd()), character.only = TRUE)
browseVignettes(package = basename(getwd()))
?checkinput::reorder_levels
```

### Update package-wide documentation

#### NEWS
Restyle and publish the `NEWS`-file at each new package release.

#### README
To provide a nice overview of functions: see
[this example](https://github.com/MJobinASU/MontyHall). After adjusting the
`README.Rmd`, `Knit` it to produce a `README.Md` file.

#### Increment package version
To increment the package version, adjust the package version in the `DESCRIPTION`
file, or run `R` as administrator and then use `usethis::use_version()`. In the
latter case, do not automatically `commit` the change, but do so manually to
adjust the commit message to something like
`Bump to version x.y.z. Breaking change: <func> no longer ...`. Indicating the
version number in the commit title makes it easier to find changes back later on.

## Performing updates

### Use GitHub Actions
To manually run GitHub Actions (GHA) set up `workflow_dispatch` (see section
`Use GitHub Actions` in the vignette *Package setup*:
`vignette("pkg_setup", package = "develcoder")`). To check if package `B`, which
depends on package `A`, is still working fine, go to the `Actions` tab of
repository `B`,
select the action you want to trigger (e.g., `R-CMD-check.yaml`), and use the
`Run workflow` button to run the GHA. You can select which branch it should
run on, but you need to trigger it once manually on the `main` branch to be
able to trigger it manually on other branches.

Scheduled jobs that failed can be rerun through the button `Re-run jobs` >
`Re-run failed jobs` > `Re-run jobs`. If the GitHub actions page (e.g.,
https://github.com/JesseAlderliesten/develcoder/actions) indicates the package
passes the R cmd check whereas the badge on the `README` indicates it fails,
make sure the R cmd check considers the `main` branch (i.e., if the `devel`
branch passes but the `main` branch fails, the badge will have `failing`): use
the `Run workflow` button to run the GHA on the `main` branch (see the previous
paragraph).

If the package declares a dependency on a specific `R` version, it is useful to
specify the minimum declared `R` version to run in addition to the ones that are
by default used in the template: add `- {os: ubuntu-latest,   r: '4.1.0'}` to
section `matrix: config:` to run `R` 4.1.0.

### pkgdown website
See the documentation about package [pkgdown](https://pkgdown.r-lib.org/) and
the [chapter](https://r-pkgs.org/website.html) from the R packages book.

``` {r pkgdown, eval = FALSE}
# To manually update the website
pkgdown::build_site()
```
Open `docs/index.html` in a webbrowser to preview the website, or look at the
files that constitute your package's website are in the local `docs/` directory.

Instead of manually updating the pkgdown website, one can use a
[GitHub Action][Use GitHub Actions] workflow (e.g.,
[pkgdown.yaml](https://github.com/JesseAlderliesten/develcoder/blob/main/.github/workflows/pkgdown.yaml))
that updates the website after a pull request or push. To set this up, run
`usethis::use_pkgdown_github_pages()`.

#### Styling
To create a thematic index instead of the default alphabetically ordered one,
see the [documentation](https://pkgdown.r-lib.org/reference/build_reference.html)
and the [example](https://github.com/r-lib/pkgdown/blob/main/pkgdown/_pkgdown.yml).

To customise the order in which `Articles` are listed, specify their order in
the `_pkgdown.yml` file (run `tools::pkgVignettes(package = "<pkg>")$names`
to get their names; see the
[documentation](https://pkgdown.r-lib.org/reference/build_articles.html) and for
example the `_pkgdown.yml`
[file](https://github.com/JesseAlderliesten/checkrpkgs/blob/master/.github/workflows/pkgdown.yaml)
and non-alphabetical order of the
[articles](https://jessealderliesten.github.io/checkrpkgs/articles/) in the
`checkrpkgs` package.

To create a custom function index, include a section `reference:` in the
`_pkgdown.yml` file with the desired headings and topics, see the
[documentation](https://pkgdown.r-lib.org/reference/build_reference.html), an
[official example](https://github.com/r-lib/pkgdown/blob/main/pkgdown/_pkgdown.yml)
and [my own example](https://github.com/JesseAlderliesten/progutils/blob/main/_pkgdown.yml).
Then build the reference with `pkgdown::build_reference()` and view it locally
by opening the file `"<pkg>\docs\index.html"` and browse to `Reference` to see
the package index.

### Merge devel-branch with master
To merge the `devel`-branch with `master`, go the the `devel`-branch on GitHub.
`<Contribute>` > `Open Pull Request`. Copy the updated `NEWS` in the
`description` field and use the button `Create pull request`.

If you see
`No conflicts with base branch: Merging can be performed automatically`, you can
use the green `Merge pull request` button, or change to `Squash and merge`.

Otherwise, you will see `This branch has conflicts that must be resolved`:
follow the instructions to resolve the conflicts. Then push the button
`Commit merge`. Then you should see `No conflicts with base branch` and you can
proceed as described in the previous paragraph.

After a successful merge, you will see a message that you can delete the
`devel`-branch, which you can do. To do so later, go to `Pull requests`, select
the `closed` tab, and scroll down to the button `Delete branch`.

### Overwrite devel-branch after merge
In `RStudio`, open the relevant project to check there are no commits left. Then
you can move to the `master` branch and `Pull` to get all updates you just
committed to the `master` branch. Then click the `New branch` button in `RStudio`
(besides the `Switch branch` icon indicating which branch (e.g., `master` or
`devel`) you are using), use `devel` as branch name and click `Create`. You will
be notified that `devel` already exists and asked if you want to overwrite it.
**If** you have just merged `devel` into `master`, you can choose to overwrite
it to have a new `devel` branch.

#### Set up a new branch
Instead of overwriting the `devel`-branch after merging into `main`, you can
also create a completely new branch:

- At the GitHub page of the package: `Branch` icon > green `New branch` button >
  specify a name for the branch > green `Create new branch` button.
- Go back to the GitHub page of the package, and at the green `Code` button >
  `copy URL to clipboard`.
- In `RStudio`: `File` > `New project` > `Version control` > `Git` and paste the
  copied URL in the `Repository URL` > `Create Project`.
- Then (still in `RStudio`) in the `Git` menu change from `master` to the
  just-created branch.

### Installing the updated package
Then you can delete the old package version from your PC (see `getwd()` for its
location) and install the updated version following your own instructions on the
GitHub pages of the relevant packages.


## Troubleshooting
To prevent `regex`-classes in example code from being interpreted as links
(which leads to the error
`@section Could not resolve link to topic ":blank:" in the dependencies or base packages`)
when running `devtools::document()`, use backticks (`` ` ``) to format a line as
code, or wrap consecutive lines in `` \code{...}` ``.

## Guidelines
- https://developer.r-project.org/, especially 
  https://developer.r-project.org/devel-guidelines.txt
- https://cran.r-project.org/web/packages/policies.html
- https://github.com/hturner/pkg-dev-ctv/blob/main/proposal.md#links-links
