---
title: "Package setup"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Package setup}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction
This vignette contains code and annotations to create a package. This code is
needed only once to set up a package from scratch. Further package development
is described in the vignette *Package development*:
`vignette("pkg_devel", package = "develcoder")`.

# Choose a name
```{r name, eval = FALSE}
pkgname <- "develcoder"

# Is the intended name valid?
stopifnot(pkgdepends::is_valid_package_name(nm = pkgname))

# Is the intended name available?
unlist(pkgdepends::pkg_name_check(name = pkgname)$basics)
utils::browseURL(url = paste0("https://github.com/search?q=", pkgname,
                              "&type=repositories"))
utils::browseURL(url = paste0("https://github.com/search?q=", pkgname,
                              "+language%3AR+&type=repositories"))
```

# Create the package
Creating the package through `usethis::create_package()` uses information stored
in `.Rprofile` files to fill in default information. Such files can exist in
various locations, for example at `D:\Userdata\<username>\Documents\.Rprofile`:
see `help("Startup")`. See the `usethis` vignette
[`usethis setup`](https://usethis.r-lib.org/articles/usethis-setup.html) and
`help(topic = "use_description", package = usethis)` for details on setting up a
package.

```{r create, eval = FALSE}
path_to_package <- file.path(getwd(), pkgname)
path_to_package
usethis::create_package(path = path_to_package)
```

Do:

- Update the `Title` field of the `DESCRIPTION` file
- Update the `Description` field of the `DESCRIPTION` file
- Check if the automatically filled in information is correct

# Add package-wide documentation

## License
A license file makes clear what use of your package is allowed. Overviews of
licenses are available at various websites:

- [Blue oak council](https://blueoakcouncil.org/list)
- [choosealicense](https://choosealicense.com/appendix/)
- [ELRA License Wizard](http://wizard.elda.org/)
- [GNU.org](https://www.gnu.org/licenses/license-list.html)
- [Open Source Initiative](https://opensource.org/licenses)
- [System Package Data Exchange](https://spdx.org/licenses/)
- [tl;drLegal](https://www.tldrlegal.com/contact-us)

Some legal details are provided at
[The Legal Side of Open Source](https://opensource.guide/legal/).

[CRAN](https://cran.r-project.org/) only
[allows](https://cran.r-project.org/web/packages/policies.html#Source-packages)
packages with licences from
[this list](https://svn.r-project.org/R/trunk/share/licenses/license.db).
`usethis` provides utility functions to add licenses (see
`help("licenses", package = "usethis")`. For example, to add a MIT license:

```{r license, eval = FALSE}
usethis::use_mit_license()
```

## Citation
A citation file makes clear how your package should be cited. See the
documentation in the `R` package [`cffr`]( https://CRAN.R-project.org/package=cffr)
and the GitHub page [`Citation file format`](https://citation-file-format.github.io/)
for details on the format.

```{r citation, eval = FALSE}
cffr::cff_write(dependencies = FALSE) # Create a citation file
```

## README
```{r readme, eval = FALSE}
usethis::use_readme_rmd()
```

Do:

- Manually update `README.Rmd`
  - Add a badge with the version number by including the following code in the
    `README.Rmd` file (the badge does **not** work if the GitHub repository is
    private):  
    `![](https://img.shields.io/github/r-package/v/<repository>/<pkg>?color=blue)`
  - See
    [`README_template.Rmd`](https://github.com/JesseAlderliesten/develcoder/blob/main/inst/templates/README_template.Rmd)
    in the folder `inst/templates`.
  - Knit the `README.Rmd` file to produce a `README.Md` file.

### Troubleshooting
The `usethis` practice of using `devtools::build_readme()` to update `README.md`
requires that `README.Rmd` and `README.md` are staged at the same time. To
remove this requirement, delete the the (hidden) file
`.git/hooks/pre-commit` from the `R` project folder (see the section
`Description` in `help("use_readme_rmd", package = "usethis")`).

## Package overview
The file `R/<pkg>-package.R` makes that `help("<pkg>")` produces a relevant
description of the package with links to GitHub and to the package website if
package `<pkg>` has been loaded (i.e., `library(<pkg>)` has been run). It is
also used by `usethis` to list imported functions, and can be configured to give
an overview of the functions in the package.

```{r overview, eval = FALSE}
usethis::use_package_doc() # Create R/<pkg>-package.R
```

## NEWS

```{r news, eval = FALSE}
usethis::use_news_md() # Create NEWS.md
```

# Set up testing infrastructure
It is convenient to have a collection of tests for your functions. Various
packages can be used to create and run such tests. To use package
[tinytest](https://CRAN.R-project.org/package=tinytest) for testing, set up its
infrastructure through:

```{r testing tinytest, eval = FALSE}
tinytest::setup_tinytest(pkgdir = ".")
devtools::document()
```

To use package [testthat](https://cran.r-project.org/package=testthat) for
testing, set up its infrastructure through:

```{r testing testthat, eval = FALSE}
usethis::use_test("<func>")
devtools::document()
```

# Set up a GitHub repository
For more information about setting up and using GitHub and Git, see the vignette
*Using Git and GitHub*: `vignette("git_github", package = "checkrpkgs")`.

```{r GitHub, eval = FALSE}
usethis::use_git(message = "Initial commit")
# If you get the error message 'You are not the current owner of the GitHub
# repository', restart R as administrator and try again.
usethis::use_github(private = TRUE)
usethis::git_vaccinate()
# See also https://forum.posit.co/t/renaming-the-default-branch/119149
usethis::git_default_branch_rename(from = "master", to = "main")
# Sometimes this takes a while before it works
usethis::git_default_branch_rediscover()
```

# Automate checks
GitHub Actions (GHA) is a continuous integration service to automatically run
code upon certain triggers. Setting GHA to run `check-no-suggests.yaml` from
[`r-lib`](https://github.com/r-lib/actions/blob/v2-branch/examples/check-standard.yaml)
checks that the code passes [`R CMD checks`](https://r-pkgs.org/R-CMD-check.html)
even if dependencies in `Suggests` are not installed.

```{r GitHub Actions, eval = FALSE}
usethis::use_github_action("check-standard")
usethis::use_github_action("check-no-suggests")
```

Then adjust the `YAML` files (i.e., `<pkg>\.github\workflows\check-standard.yaml`
and `<pkg>\.github\workflows\check-no-suggests.yaml`) to include some other
useful triggers for GHAs (see the template files `check-standard.yaml` and
`check-no-suggests.yaml` in the folder `inst/templates`):

- you made changes, or someone else proposed changes to code in the current
  repository:  
  add `pull_request:` to section `on:` to run GHA on pull requests.
  NB. You can remove `push` from section `on:`: it should be sufficient to run R
  CMD check locally before pushing and run GHA on pull requests.
- you want to check if a reverse dependency (i.e., a package that depends on a
  package you changed) is still working fine:  
  add `workflow_dispatch:` to section `on:` in the YAML file of the reverse
  dependency to be able to manually trigger GHA. See section
  `Use GitHub Actions` in `pkg_devel.Rmd` how to use it.
- someone else made changes to packages your package depends on:  
  add `schedule: - cron: "23 4 * * 6"` to section `on:` to run every Saturday on
  04:23 UTC. The cron specification consists of five elements that indicate the
  minute (0 - 59), hour (0 - 23), day of the month (1 - 31), month (1 - 12), and
  day of the week (0 - 6). This timing is approximate and depends on how busy
  the servers are.
- If the package declares a dependency on a minimum `R` version, it is useful
  to specify the minimum declared `R` version to run in addition to the ones
  that are by default used in the template: for example, add  
  `- {os: ubuntu-latest, r: '4.1.0'}` to section `matrix: config:` to run
  `R` version 4.1.0 if your package declares `R` 4.1.0 as minimum version.

For further documentation, see section `GHA: documentation and help` in the
vignette *Package development*: `vignette("pkg_devel", package = "develcoder")`.

# Set up a package website
One advantage of using a website for your package is that it automatically
creates [`hyperlinked cross-references`](https://r-pkgs.org/vignettes.html#links)
to functions and documentation you refer to.

To set up a package website, you can use `usethis::use_pkgdown()`. To also set
up a [GitHub action][Automate checks] to automatically build and update your
site if you push changes to your package to GitHub (defined in the file
`<pkg>\.github\workflows\pkgdown.yaml`), run the following code:

```{r pkgdown, eval = FALSE}
# calls usethis::use_pkgdown() and creates '<pkg>\.github\workflows\pkgdown.yaml'
usethis::use_pkgdown_github_pages()
devtools::document() # To update package-wide documentation
```

After a few minutes, the website should be present. The URL has the structure
`https://<username>.github.io/<repository>/`, e.g.,
`https://jessealderliesten.github.io/develcoder/`.

Refer from the `README` to the pkgdown website and add the line
`light-switch: true` below the `template` heading in the `_pkgdown.YML` file
(located in the top-level package folder or in folder `pkgdown`; can also have
extension `.YAML`; it is **not** the file `<pkg>\.github\workflows\pkgdown.yaml`
that defines the [GitHub Action][Automate checks] to automatically build and
update your site if you push changes to your package to GitHub) to add a
light-switch to the pkgdown-website (see the
[documentation](https://pkgdown.r-lib.org/articles/customise.html#light-switch)
and the
[example code](https://github.com/JesseAlderliesten/develcoder/blob/main/_pkgdown.yml)
for this repository).

See the chapter about [`pkgdown`](https://r-pkgs.org/website.html) in the book
[`R packages`](https://r-pkgs.org/) and the documentation for package
[`pkgdown`](https://pkgdown.r-lib.org/) for further details.

# Further development
Further package development is described in the vignette *Package development*:
`vignette("pkg_devel", package = "develcoder")`.
