---
title: "Package setup"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Package setup}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction
This file contains code and annotations that are useful when creating a package.

## Choose a name
```{r name, eval = FALSE}
pkgname <- "develcoder"

# Is the intended name valid?
stopifnot(pkgdepends::is_valid_package_name(nm = pkgname))

# Is the intended name available?
unlist(pkgdepends::pkg_name_check(name = pkgname)$basics)
browseURL(url = paste0("https://github.com/search?q=", pkgname, "&type=repositories"))
browseURL(url = paste0("https://github.com/search?q=", pkgname, "+language%3AR+&type=repositories"))
```

## Create the package
This uses information stored in your `.Rprofile` file (located at
`D:\Userdata\Jesse Nieuw\Documents\.Rprofile`, see `help(Startup)` for details),
see `usethis::use_description()` and the
[usethis vignette](https://usethis.r-lib.org/articles/usethis-setup.html).

```{r create, eval = FALSE}
path_to_package <- file.path("D:", "Userdata", "Jesse Nieuw", "Documents", "R", pkgname)
usethis::create_package(path = path_to_package)
```

Do:

- Update the `Title` field of the `DESCRIPTION` file.
- Update the `Description` field of the `DESCRIPTION` file.

## Add package-wide documentation
A license file makes clear what use of your package is allowed. See
[choosealicense](https://choosealicense.com/) and the
[GNU FAQ](https://www.gnu.org/licenses/gpl-faq.html) for overviews of licenses and
[this blogpost](https://writing.kemitchell.com/2016/09/21/MIT-License-Line-by-Line)
for details on the MIT License.

A citation file makes clear how your package should be cited, see the
[Citation file format GitHub page](https://citation-file-format.github.io/) for
details on the format.

`R/<pkgname>-package.R` is used by `usethis` to list imported functions and can
also be used as help-page such that `help(<pkgname>)` produces a description of
the package with an overview of its functions.

```{r license, eval = FALSE}
# Run R as administrator
usethis::use_mit_license() # Create license files for a permissive license
cffr::cff_write() # Create a citation file
usethis::use_package_doc() # Create R/<pkgname>-package.R
usethis::use_readme_Rmd() # Create README.Rmd
usethis::use_news_md() # Create NEWS.md
```

Do:

- Keep `license.Rmd`: it is required by `R CMD check`.
- Manually update `README.Rmd`
  - See the template file `README_template.Rmd` in the folder `templates`.
  - Add a badge with the version number by including the following code in the
    `README.Rmd` file (replace `<pkgname>` with the actual package name; the
    badge does not work if the GitHub repository is private):
    `![](https://img.shields.io/github/r-package/v/JesseAlderliesten/<pkgname>?color=blue)`
  - `Knit` the `README.Rmd` file to produce a `README.Md` file.
- Update the `NEWS`-file (see the template file `README_template.Rmd` in the
  folder `templates`).

NB. I do *not* follow the `usethis` practise of using `devtools::build_readme()`
to update `README.md` because that imposes the requirement to have `README.Rmd`
and `README.md` staged at the same time. To remove this requirement if it was
accidentally introduced, delete the the (hidden) file `.git/hooks/pre-commit`
from the R-project folder (see the `Description` section in
`help(devtools::build_readme())`.

## Importing packages or functions
Specifying a minimum package version when importing packages ensures that used
features are actually present. `min_version = TRUE` uses the currently installed
package version. Alternatively, check which features are used and in which
version those were introduced to find a minimum version. That can also be done
afterwards by specifying minimum versions in checks through GitHub actions.

```{r dependencies, eval = FALSE}
usethis::use_package(package = "checkinput", type = "Imports", min_version = TRUE)
usethis::use_dev_package(package = "checkinput", type = "Imports",
                         remote = "github::JesseAlderliesten/checkinput")

usethis::use_package(package = "progutils", type = "Imports", min_version = TRUE)
usethis::use_dev_package(package = "progutils", type = "Imports",
                         remote = "github::JesseAlderliesten/progutils")
devtools::document()
```


## Set up testing
```{r testing, eval = FALSE}
tinytest::setup_tinytest(pkgdir = ".")
usethis::use_package(package = "tinytest", type = "Suggests")
devtools::document()
```

## Set up a GitHub repository
For more information about setting up and using GitHub and Git, see
`vignette(topic = "git_github", package = "checkrpkgs")`.

```{r GitHub, eval = FALSE}
usethis::use_git(message = "Initial commit")
# If you get the error message you are not the current owner of the GitHub
# repository, restart R as administrator and try again.
usethis::use_github(private = TRUE)
```

## Use GitHub Actions
GitHub Actions (GHA) is a continuous integration service to automatically run
code upon certain triggers. Setting GHA to run `check-no-suggests.yaml` from
[r-lib](https://github.com/r-lib/actions/blob/v2-branch/examples/check-standard.yaml)
ensures the code passes `R-CMD check` when dependencies in `Suggests` are not
installed.

```{r GitHub Actions, eval = FALSE}
usethis::use_github_action("check-standard")
usethis::use_github_action("check-no-suggests")
```

Then adjust the YAML file (i.e., `<pkg>\.github\workflows\check-no-suggests.yaml`)
to include some other useful triggers for GHAs (see the template file
`check-no-suggests.yaml` in the folder `templates`):

- you made changes to code in the current repository: add `'push:'` to section
  `'on:'` to run GHA on pushes.
- someone else proposed changes to code in the current repository: add
  `'pull_request:'` to section `'on:'` to run GHA on pull requests.
- you changed package `A` and want to check if package `B`, which depends on
  package `A`, is still working fine): add `'workflow_dispatch:'` to section
  `'on:'` to be able to manually trigger GHA. See section `Use GitHub Actions`
  in `pkg_devel.Rmd` on how to use it.
- someone else made changes to packages your package depends on: add e.g.
  `'schedule: - cron: "23 4 * * 6"'` to section `'on:'` to run every Saturday on
  04:23 UTC. The cron specification consists of five elements that indicate the
  minute (0 - 59), hour (0 - 23), day of the month (1 - 31), month (1 - 12), and
  day of the week (0 - 6).
- If the package declares a dependency on a specific `R` version, it is useful
  to specify the minimum declared `R` version to run in addition to the ones
  that are by default used in the template: add e.g.,
  `'- {os: ubuntu-latest,   r: '4.1.0'}'` to section `'matrix: config:'` to run
  `R` version 4.1.0.

### Further documentation
See the section about
[GitHub Actions](https://r-pkgs.org/software-development-practices.html#sec-sw-dev-practices-gha)
in [R packages](https://r-pkgs.org/) and the GitHub documentation about
[workflow syntax](https://docs.github.com/en/actions/reference/workflows-and-actions/workflow-syntax)

## Using pkgdown
See the chapter about [pkgdown](https://r-pkgs.org/website.html) from the book
[R packages](https://r-pkgs.org/) and the documentation for package
[pkgdown](https://pkgdown.r-lib.org/).

```{r pkgdown, eval = FALSE}
usethis::use_pkgdown_github_pages() # This first calls usethis::use_pkgdown()
```

After a few minutes, the website should be present. The URL has the structure
`https://USERNAME.github.io/REPONAME/`, e.g.,
`https://jessealderliesten.github.io/checkrpkgs/`.
