---
title: "Package development"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Package development}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction
This vignette contains code and annotations that are useful when developing a
package, including when preparing to release a new package version and setting
up for a new version. It assumes the package has already been set up, as
described in the vignette *Package setup*:
`vignette("pkg_setup", package = "develcoder")`.


# Adding functions
## Set up
```{r functions, eval = FALSE}
usethis::use_r("<func>")
devtools::document() # also runs devtools::load_all()
?<func>() # view help-page of the function
```

Do:

- Manually update the `NEWS` file: ``- Add `<func>()` to ...``.
- When adding new function arguments to an existing function, it is best
  practice to place the new argument after existing arguments and use a default
  value that matches the old behaviour: then code written for the old version
  will work the same as before the change, even if it relies on positional
  matching, such that it is a **non**-breaking change.
- When **renaming** functions, remember to also change their names in the
  `_pkgdown.yml` file if that file is used to create a custom index for the
  website (`pkgdown::build_reference()` warns about it if you forget).

## Documentation
```{r documentation, eval = FALSE}
#' Title
#'
#' Description
#'
#' # Since `roxygen2` 8.0.0, it is possible to specify which arguments to inherit
#' @inheritParams is_logical allow_NA # for a function in the current package
#' @inheritParams utils::installed.packages # for a function in another package
#' @param x Vector of names to test.
#' @param allow_NA `TRUE` or `FALSE`: allow [NA]s of the correct type in `x`?
#' @param x,y Separate arguments by a comma without a space to create a single
#' description for multiple arguments.
#'
#' @details
#'
#' @returns
#' `TRUE` or `FALSE`, returned [invisibly][invisible].
#'
#' @section Side effects:
#' The directory indicated by the returned path is
#' [created][progutils::create_dir()] if it does not yet exist.
#'
#' @section Notes:
#' Some notes.
#'
#' @section Programming notes:
#' Some programming notes. Create a GitHub issue instead of To do / Wishlist
#' sections.
#'
#' @seealso
#' Section `Details` of [make.names()] and the [\R FAQ about valid names](
#' https://CRAN.R-project.org/doc/manuals/R-FAQ.html#What-are-valid-names_003f)
#' on the syntactical validity of names.
#'
#' @family collections of checks on type and length
#'
#' @examples
#' all_names(x = c("a", "b2a")) # TRUE
#'
#' @export
func_name <- function(x, allow_underscores = TRUE) {
  stopifnot(checkinput::is_logical(allow_underscores))
  ...
}
```

### Inherit sections
Inherited sections are **silently** ignored if that section is also defined in
the file itself.

``` {r inherit docs, eval = FALSE}
#' @inherit is_number return
#' @inherit is_number details
#' @inheritSection is_logical Programming notes
#' @inheritSection is_logical @note
```

### Add examples
Examples should **not** write to the working directory (e.g., using `getwd()`)
but to a temporary directory that is cleaned up afterwards. See the section
`Usage in practice` in `help("create_tempdir", package = "progutils")`.

Examples rendered on a website created with [pkgdown](https://pkgdown.r-lib.org/)
will display the output underneath the code. However, help-files accessed
through `help(<func>)` will not, such that it might be useful to include a short
comment regarding what feature of the output is notable.

``` {r examples, eval = FALSE}
devtools::run_examples()
```

## Add tests
Tests should **not** write to the working directory (e.g., using `getwd()`) but
to a temporary directory that is cleaned up afterwards. See the section
`Usage in practice` in `help("create_tempdir", package = "progutils")`. More
generally, tests should not make changes (without restoring the original state)
that could influence subsequent tests. Examples of such changes are any settings
and changes to the state of the global environment (see the vignette
[test fixtures](https://testthat.r-lib.org/articles/test-fixtures.html) from package `testthat` for more details).

``` {r tests, eval = FALSE}
tinytest::test_all() # Run all tests of a package
# Run specific file, selecting the file by name
tinytest::run_test_file(file.path(getwd(), "inst", "tinytest", "test_funcname.R"))
# Run specific file, selecting the file by order
tinytest::run_test_file(
  list.files(file.path(getwd(), "inst", "tinytest"), full.names = TRUE)[1L])

# expect_silent(...) tests that no warnings or errors are emitted, NOT that no
# messages are emitted.
tinytest::expect_silent(expect_true(func_name(x = x, arg = arg)))
tinytest::expect_warning(
  expect_equal(func_name(x = "a", arg = arg), 3),
  pattern = "...", strict = TRUE, fixed = TRUE)
tinytest::expect_error(func_name(x = "a", arg = arg),
             pattern = "is_number(x) is not TRUE", fixed = TRUE)
```

Testing file paths requires some thought because the file separator depends on
the operating system (see `.Platform$file.sep`) and the backward slash is used
as escape character in R such that it needs to be escaped itself by doubling
them. Thus, a check on the presence of successive slashes and backslashes in
[string][is_character()] `string` would use
`grepl(pattern = "//", x = string, fixed = TRUE)` and
`grepl(pattern = "\\\\", x = string, fixed = TRUE)`. The message to point out
their presence would be written as `message("Successive '/' or '\\\\'")` which
would be printed as `Successive '/' or '\\'`.

To circumvent the hassle of getting the correct type and number of slashes to
compare with the path recorded in a message, check only for fixed parts of the
message (e.g., `"Repeated"`), possibly followed by a check like
`tinytest::expect_true(fs::dir_exists(string))`.

## Update dependencies
Specifying a minimum package version when importing packages ensures that used
features are actually present. Setting `min_version = TRUE` in 
`usethis::use_package()` uses the currently installed package version.
Alternatively, check which features are used and in which version those were
introduced to find a minimum version. That can also be done by specifying
minimum versions in checks through GitHub actions.

Since R 4.6.0, `read.dcf()` recognises lines starting with `#` as comment lines,
making it possible to indicate in comments in the `DESCRIPTION` file why a
particular package version is needed.

``` {r dependencies, eval = FALSE}
# Declare a minimum version for R itself
usethis::use_package("R", type = "Depends", min_version = "4.1.0")

# Declare a dependency on a package: 'use_package()' to declare a minimum
# version and 'use_dev_package()' to specify the remote repository to download
# the package from.
usethis::use_package(package = "checkinput", type = "Imports", min_version = TRUE)
usethis::use_dev_package(package = "checkinput", type = "Imports",
                         remote = "github::JesseAlderliesten/checkinput")
usethis::use_package(package = "progutils", type = "Imports", min_version = TRUE)
usethis::use_dev_package(package = "progutils", type = "Imports",
                         remote = "github::JesseAlderliesten/progutils")
devtools::document()
```

Do:

- Specify the minimum declared `R` version as workflow in GitHub Actions, see
the section [Use GitHub Actions].

See [xfun](https://github.com/yihui/xfun/blob/main/R/zzz.R) how to implement
using a function that is present in 'new' versions of R also in older versions
of R. The [cross](https://github.com/DavisVaughan/cross) package can be used to
benchmark old and new versions.

### Importing packages or functions
The standard way to use functions from other packages is to import the package
(i.e., put it in the `Imports:` field of the `DESCRIPTION` file) and in code use
the package name followed by two colons and the function name, e.g.,
`utils::osVersion()`. If that is not possible, for example because the function
is an operator, it is necessary to also list the function in the `NAMESPACE`
file, e.g., to add the line `#' @importFrom utils osVersion` to the file
`R/<pkg>-package.R` that was created by `usethis::use_package_doc()`. This
is done by `usethis::use_import_from(package = "utils", fun = "osVersion")`.

``` {r imports, eval = FALSE}
usethis::use_import_from(package = "utils", fun = "osVersion")
usethis::use_package("utils", type = "Suggests", min_version = "4.1.0")
```


# Adding vignettes
## Set up
``` {r vignettes, eval = FALSE}
usethis::use_vignette("my_vignette", title = "Some title")
# Add suggested dependencies on 'knitr' and 'rmarkdown'
usethis::use_package(package = "knitr", type = "Suggests")
usethis::use_package(package = "rmarkdown", type = "Suggests")
devtools::document()
pkgdown::build_article()
browseVignettes(package = basename(getwd()))
```

If no vignettes are visible, `devtools::install()` was probably run with the
default argument `build_vignettes = FALSE`, use `build_vignettes = TRUE` instead:
`devtools::install(quick = FALSE, upgrade = FALSE, build_vignettes = TRUE)`. Run
`tools::pkgVignettes(package = "<pkg>")` for information about which
vignettes R finds.

Use `utils::vignette("<vignette_title>")` (e.g., `utils::vignette("pkg_devel")`)
and `tools::Rd2txt("path/to/file.Rd")` to render a vignette or an `Rd`-file in
the help-pane of RStudio, respectively.

## Styling
See the vignette *RMarkdown and knitr*:
`vignette("rmarkdown_knitr", package = "develcoder")`) on using RMarkdown to
style vignettes.

Put the next lines in the header of vignettes to get a table of contents:
``` {r vignette style, eval = FALSE}
output:
  rmarkdown::html_vignette:
  toc: true
toc_depth: 3
```

## Linking
For internal links to other sections in the same document ('Section', 'above',
'below'): `[<Section title>]` or `[<link text>][<Section title>]`.

Linking to a help page of a package: `help("<func>", package = "<pkg>")`
(`help("<func>::<pkg>")` does **not** work).

There is [**no**](https://r-pkgs.org/vignettes.html#links) official way to link
from vignettes to help-pages. Using relative paths (e.g.,
`[<func>](../help/<func>)`, as proposed in a
[StackOverflow answer](https://stackoverflow.com/questions/34946219)) is
impaired by the change in the location of files when the package is installed.
In addition, linking on the `pkgdown` website does **not** work in that format
(`pkgdown` recognises calls like `` `<func>()` `` and
`` `<pkg>::<func>()` ``, see the
[documentation](https://pkgdown.r-lib.org/articles/linking.html)). For a more
advanced way of linking from vignettes to help-pages that might work, see
[here](https://github.com/dmurdoch/rgl/commit/bbc84447c2a6efed42907fbac176e9569b868d8f).

To link from help-pages to vignettes, use
``` {r link vignettes, eval = FALSE}
The vignette *<vignette title>*: `vignette("<vignette>", package = "<pkg>")` 
```


# Adding miscellaneous files
Non-standard files or folders should be added in the `inst/` directory to pass
[R CMD checks](https://r-pkgs.org/R-CMD-check.html). Those files and folders
will be in the top directory in the installed package.


# Preparing for updates

## Check tests
Check for which functions no test file has been written.

``` {r update tests, eval = FALSE}
devtools::document()
tinytest::test_all()

curr_pkg_name <- basename(getwd())
files_R <- list.files(file.path(".", "R"))
files_man <- list.files(path = file.path(".", "man"), pattern = "\\.Rd$")
proj_pkg_R <- paste0(curr_pkg_name, "-package.R")
if(proj_pkg_R %in% files_R) {
  files_R <- files_R[files_R != proj_pkg_R]
  files_man <- files_man[files_man != paste0(proj_pkg_R, "d")]
}
if("reexports.Rd" %in% files_man) {
  files_man <- files_man[files_man != "reexports.Rd"]
  # Based on https://stackoverflow.com/a/74487073/32365738
  reexported_funcs <- progutils::not_in(getNamespaceExports(curr_pkg_name),
                                        ls(envir = asNamespace(curr_pkg_name)))
  if(length(reexported_funcs) == 0L) {
    warning("File 'reexports.Rd' is present but no re-exported functions found!")
  } else {
    files_R <- progutils::not_in(files_R, paste0(reexported_funcs, ".R"))
  }
}
expected_man <- sub(pattern = "\\.Rd$", replacement = ".R", x = files_man)
expected_files <- sort(unique(paste0("test_", c(files_R, expected_man))))
tests_missing <- expected_files[!(expected_files %in% list.files("./inst/tinytest"))]
if(length(tests_missing) > 0L) {
  warning("No test-file found for file ", progutils::paste_quoted(tests_missing))
}
```

## Automated checks
`tools::CRAN_check_results()`, `tools::CRAN_check_details()` and
`tools::CRAN_check_issues()` give information about the current check status of
CRAN packages.

``` {r checks, eval = FALSE}
devtools::document()
# See https://r-pkgs.org/R-CMD-check.html on how to proceed if checks fail. If
# checks fail because of `latex` errors when building the manual, you can use
# `manual = FALSE`.
devtools::check(force_suggests = TRUE, run_dont_test = TRUE)
devtools::release_checks()
if(utils::packageVersion("devtools") >= "2.5.0") {
  # check for missing `\value` and `\examples` fields in `Rd` files
  devtools::check_doc_fields()
}

tools::check_package_urls(dir = getwd())
tools::check_package_dois(dir = getwd())

pkg_env <- devtools::load_all()$env
codetools::checkUsageEnv(env = pkg_env, all = TRUE, suppressParamAssigns = TRUE)

# - 'pkgcheck' depends on 'pkgstats', which needs the system libraries
#   'ctags-universal' and 'GNU global'. Run pkgstats::ctags_test() to check if
#   those are installed. If not, you might have to run pkgstats::ctags_install();
#   see also the output of library(pkgstats) and the documentation at
#   https://docs.ropensci.org/pkgstats/articles/installation.html
# - pkgcheck::pkgcheck() and goodpractice::gp() have quite some overlap. See
#   https://github.com/ropensci-review-tools/goodpractice/issues/284 
# - The lintr issue `Use == instead of %in% for scalar comparison` should be
#   ignored for 'x %in% y' if 'x' or 'y' might contain 'NA' that should be
#   removed: 'x %in% NA' and 'NA %in% y' return 'FALSE', whereas 'x == NA' and
#   'NA == y' return 'NA'.
res_pkgcheck <- pkgcheck::pkgcheck()
res_pkgcheck
res_goodpractice <- goodpractice::gp()
res_goodpractice

# Check spelling (requires system library like Aspell, see http://aspell.net/)
utils::aspell_package_Rd_files(dir = getwd())
utils::aspell_package_vignettes(dir = getwd())
```

## Locally test update
``` {r update, eval = FALSE}
devtools::document()
.libPaths() # Check if output of .libPaths() is correct.
devtools::install(quick = FALSE, upgrade = FALSE, build_vignettes = TRUE)

# Load the package and view the help files as usual outside devtools:
library(basename(getwd()), character.only = TRUE)
browseVignettes(package = basename(getwd()))
?progutils::reorder_levels
```

## Check reverse dependencies
**Acknowledgement**
This section reproduces a
[contribution](https://stat.ethz.ch/pipermail/r-package-devel/2026q2/012397.html)
by Ivan Krylov to the
[R-pkg-devel mailing list](https://stat.ethz.ch/mailman/listinfo/r-package-devel)
of 26 May 2026.

To check if packages that use your package still pass their checks if you made
breaking changes, install the new version of your package, run
`tools::package_dependencies(reverse = TRUE, which = "most", recursive = "strong")`
to get the list of reverse dependencies, then use `utils::download.packages(...)`
to obtain their latest tarballs and finally run `tools::check_packages_in_dir()`
to check them. See also `xfun::rev_check()` and package
[revdepcheck](https://revdepcheck.r-lib.org/).

## Update package-wide documentation

- `README`: after adjusting the `README.Rmd`, `Knit` it to produce a `README.Md`
  file. You can follow the `usethis` practice of using `devtools::build_readme()`
  to update `README.md`, which requires that `README.Rmd` and `README.md` are
  staged at the same time. To remove this requirement if you do not want it,
  delete the the (hidden) file `.git/hooks/pre-commit` from the R-project folder
  (see the `Description` section in `help("use_readme_rmd", package = "usethis")`).
- `NEWS`: restyle and publish the `NEWS`-file at each new package release.
- `DESCRIPTION`: increment the package version in the `DESCRIPTION` file, or run
  `R` as administrator and then use `usethis::use_version()`. In the latter
  case, do not automatically `commit` the change, but do so manually to adjust
  the commit message to `Bump to version x.y.z`. Indicating the version number
  in the commit title makes it easier to find changes back later on (the pull
  request can have a description of the changes, i.e., the updated section of
  the `NEWS` file).
- `CITATION.cff`: update your citation file to reflect the new package version:

  ```{r citation, eval = FALSE}
  cffr::cff_write()
  ```

# Updating

## Use GitHub Actions
To manually run GitHub Actions (GHA) set up `workflow_dispatch` (see section
`Automate checks` in the vignette *Package setup*:
`vignette("pkg_setup", package = "develcoder")`). To check if package `B`, which
depends on package `A`, is still working fine, go to the `Actions` tab of
repository `B`,
select the action you want to trigger (e.g., `R-CMD-check.yaml`), and use the
`Run workflow` button to run the GHA. You can select which branch it should
run on, but you need to trigger it once manually on the `main` branch to be
able to trigger it manually on other branches.

Scheduled jobs that failed can be rerun through the button `Re-run jobs` >
`Re-run failed jobs` > `Re-run jobs`. If the GitHub actions page (e.g.,
https://github.com/JesseAlderliesten/develcoder/actions) indicates the package
passes the R cmd check whereas the badge on the `README` indicates it fails,
make sure the R cmd check considers the `main` branch (i.e., if the `devel`
branch passes but the `main` branch fails, the badge will have `failing`): use
the `Run workflow` button to run the GHA on the `main` branch (see the previous
paragraph).

If the package declares a dependency on a minimum `R` version, it is useful to
specify the minimum declared `R` version to run in addition to the ones that are
by default used in the template: add `- {os: ubuntu-latest,   r: '4.1.0'}` to
section `matrix: config:` to run `R` 4.1.0.

## pkgdown website
See the documentation about package [pkgdown](https://pkgdown.r-lib.org/) and
the [chapter](https://r-pkgs.org/website.html) from the R packages book.

``` {r pkgdown, eval = FALSE}
# To manually update the website
pkgdown::build_site()
```
Open `docs/index.html` in a web browser to preview the website, or look at the
files that constitute your package's website are in the local `docs/` directory.

Instead of manually updating the pkgdown website, one can use a
[GitHub Action][Use GitHub Actions] workflow (e.g.,
[pkgdown.yaml](https://github.com/JesseAlderliesten/develcoder/blob/main/.github/workflows/pkgdown.yaml))
that updates the website after a pull request or push. To set this up, run
`usethis::use_pkgdown_github_pages()`.

### Styling
To create a thematic index instead of the default alphabetically ordered one,
see the [documentation](https://pkgdown.r-lib.org/reference/build_reference.html)
and the [example](https://github.com/r-lib/pkgdown/blob/main/pkgdown/_pkgdown.yml).

To customise the order in which `Articles` are listed, specify their order in
the `_pkgdown.yml` file (run `tools::pkgVignettes(package = "<pkg>")$names`
to get their names; see the
[documentation](https://pkgdown.r-lib.org/reference/build_articles.html) and for
example the `_pkgdown.yml`
[file](https://github.com/JesseAlderliesten/checkrpkgs/blob/master/.github/workflows/pkgdown.yaml)
and non-alphabetical order of the
[articles](https://jessealderliesten.github.io/checkrpkgs/articles/) in the
`checkrpkgs` package.

To create a custom function index, include a section `reference:` in the
`_pkgdown.yml` file with the desired headings and topics, see the
[documentation](https://pkgdown.r-lib.org/reference/build_reference.html), an
[official example](https://github.com/r-lib/pkgdown/blob/main/pkgdown/_pkgdown.yml)
and [my own example](https://github.com/JesseAlderliesten/progutils/blob/main/_pkgdown.yml).
Then build the reference with `pkgdown::build_reference()` and view it locally
by opening the file `"<pkg>\docs\index.html"` and browse to `Reference` to see
the package index.

## Merge devel-branch with master
To merge the `devel`-branch with `master`, go the the `devel`-branch on GitHub.
`<Contribute>` > `Open Pull Request`. Copy the updated `NEWS` in the
`description` field and use the button `Create pull request`.

If the message
`No conflicts with base branch: Merging can be performed automatically`, is
displayed, you can use the green `Merge pull request` button, or change to
`Squash and merge`.

Otherwise, the message `This branch has conflicts that must be resolved` will be
displayed: follow the instructions to resolve the conflicts. Then push the
button `Commit merge`. Then you should see `No conflicts with base branch` and
you can proceed as described in the previous paragraph.

After a successful merge, you will see a message that you can delete the
`devel`-branch, which you can do. To do so later, go to `Pull requests`, select
the `closed` tab, and scroll down to the button `Delete branch`.

### Overwrite devel-branch after merge
In `RStudio`, open the relevant project to check there are no commits left. Then
you can move to the `master` branch and `Pull` to get all updates you just
committed to the `master` branch. Then click the `New branch` button in `RStudio`
(besides the `Switch branch` icon indicating which branch (e.g., `master` or
`devel`) you are using), use `devel` as branch name and click `Create`. You will
be notified that `devel` already exists and asked if you want to overwrite it.
**If** you have just merged `devel` into `master`, you can choose to overwrite
it to have a new `devel` branch.

### Set up a new branch
Instead of overwriting the `devel`-branch after merging into `main`, you can
also create a completely new branch:

- At the GitHub page of the package: `Branch` icon > green `New branch` button >
  specify a name for the branch > green `Create new branch` button.
- Go back to the GitHub page of the package, and at the green `Code` button >
  `copy URL to clipboard`.
- In `RStudio`: `File` > `New project` > `Version control` > `Git` and paste the
  copied URL in the `Repository URL` > `Create Project`.
- Then (still in `RStudio`) in the `Git` menu change from `master` to the
  just-created branch.

## Installing the updated package
Then you can delete the old package version from your PC (`find.package("<pkg>")`
gives its location) and install the updated version following your own
instructions on the GitHub pages of the relevant packages.


# Troubleshooting
To prevent `regex`-classes in example code from being interpreted as links
(which leads to the error
`@section Could not resolve link to topic ":blank:" in the dependencies or base packages`)
when running `devtools::document()`, use backticks (`` ` ``) to format a line as
code, or wrap consecutive lines in `` \code{...}` ``.

# Documentation and help

## Guidelines on package development

- The
  [Writing R extensions](https://cran.r-project.org/doc/manuals/r-release/R-exts.html)
  manual is the official guide on extending R
- The [CRAN Repository policy](https://cran.r-project.org/web/packages/policies.html)
  and [BioConductor package guide](https://contributions.bioconductor.org/) are
  the official guides for [CRAN](https://cran.r-project.org/) and
  [BioConductor](https://bioconductor.org/)
- [EpiverseTrace](https://epiverse-trace.github.io/) has
  [blueprints for software development](https://epiverse-trace.github.io/blueprints/)
- The [R Contribution Working Group](https://contributor.r-project.org/) has an
  [R development guide](https://contributor.r-project.org/rdevguide/)
- [rOpenSci](https://ropensci.org/) has a
  [package development guide](https://devguide.ropensci.org/), a
  [Statistical Software Peer Review](https://stats-devguide.ropensci.org/) guide
  and a useful section [Publish packages](https://docs.r-universe.dev/publish.html)
  in the [R universe documentation](https://docs.r-universe.dev/)

## Other useful resources

- The book [R packages](https://r-pkgs.org/) by Wickham and Bryan
- The book [R in production](https://r-in-production.org/) by Wickham and Masiello
- The chapter
  [Building R Packages](https://bookdown.org/rdpeng/RProgDA/building-r-packages.html)
  of [Mastering Software Development in R](https://bookdown.org/rdpeng/RProgDA/)
  by Peng, Kross, and Anderson
- The website [What They Forgot to Teach You About R](https://rstats.wtf/)
- The R [developer page](https://developer.r-project.org/) contains
  [developer guidelines](https://developer.r-project.org/devel-guidelines.txt)
  and [technical documentation](https://developer.r-project.org/TechDocs/)
- [Posit](https://posit.co/) has a
  [cheatsheet](https://opensource.posit.co/resources/cheatsheets/package-development/)
  about package development
- [Package Development and Maintenance](https://github.com/hturner/pkg-dev-ctv/blob/main/proposal.md)
  (currently a proposal for a
  [CRAN Task View](https://cran.r-project.org/web/views/)) provides an annotated
  thematic collection of R packages that assist in developing R packages
- The separate R [NEWS](https://cran.r-project.org/doc/manuals/r-devel/NEWS.html)
  for the development branch
