---
title: "R Markdown and knitr"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{R Markdown and knitr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction
This vignette contains information about using
[R Markdown](https://pkgs.rstudio.com/rmarkdown/) and
[knitr](https://yihui.org/knitr/).

An R Markdown file is an R script with markup to enable combining the script
(including the comments) with its output and additional text in a single
document. The R package [rmarkdown](https://cran.r-project.org/package=rmarkdown)
and either [RStudio](https://posit.co/products/open-source/rstudio) or
[Pandoc](https://pandoc.org/) are required to create an R Markdown file.

To generate a PDF as output, a [LaTeX](https://www.latex-project.org/get/)
distribution needs to be installed. See the 'R markdown cookbook' and the
'Definitive guide' (both mentioned at [Useful resources] below) for extensive
help. When installing [MikTex](https://miktex.org/) as LaTeX distribution, set
`install for all users` and `always install missing packages on the fly` (for
all users), see
[this Stack Exchange post](https://tex.stackexchange.com/questions/27138/). An
alternative to MikTeX is [TinyTeX](https://yihui.org/tinytex/) (for debugging
see [here](https://yihui.org/tinytex/r/)):

```{r tinytex, eval = FALSE}
writeLines(c(
  '\\documentclass{article}',
  '\\begin{document}', 'Hello world!', '\\end{document}'
), 'test.tex')
tinytex::pdflatex('test.tex')
options(tinytex.verbose = FALSE)
```

## Useful resources
- the [R markdown cookbook](https://pkg.yihui.org/rmarkdown-cookbook/) by
  Y. Xie, C. Dervieux, and E. Riederer
- the [Definitive guide](https://pkg.yihui.org/rmarkdown-book/) by Y. Xie, J.J.
  Allaire, and G. Grolemund
- a [cheatsheet](https://opensource.posit.co/resources/cheatsheets/rmarkdown/)
  and [website](https://rmarkdown.rstudio.com/) on R Markdown from
  [Posit](https://posit.co/)
- the maintainer's website on [knitr](https://yihui.org/knitr/)
- the [Markdown guide](https://www.markdownguide.org/)

## Global Settings for knitr used by R Markdown
For a list of available options, see `str(knitr::opts_chunk$get())` and the
overviews at the maintainer's website ([here](https://yihui.org/knitr/options/)
and [here](https://yihui.org/knitr/objects/)).

The option `echo = FALSE` hides code from the output. Use numeric values to
include the code of particular chunks. To collect all code as an the appendix
at the end of of a document, see the section about putting
[code in an appendix](https://pkg.yihui.org/rmarkdown-cookbook/code-appendix)
from the 'rmarkdown cookbook'.

If the option `include = FALSE` is used, code and results are not included in
output, but code **is** executed such that its results can be used in other code
chunks (useful as option for particular chunks, not as global option).

If the option `error = TRUE` is used, code execution will **not** stop on error
(unless `include = FALSE`), but instead the error message is included in the
output. This is useful in non-interactive use in production settings.
If the option `error = FALSE` is used, code execution stops on error, which
makes more sense in interactive use during development of the script.


## Loading files
Files that have to be used by the R Markdown script (e.g., R scripts that are
sourced, data files that are read) should be placed in the same directory as the
R Markdown file, because the working directory when evaluating R code chunks is
the directory of the input document by default. The working directory can be
changed using `opts_knit$set(root.dir = ...)` but should **not** be changed using
`setwd()`, see the `Note` in `help("knit", package = "knitr")`.

An inferior alternative to including `file = ...` in the header is including
```{r source, eval = FALSE}
source("<path/to/file>.R", local = knitr::knit_global(),
       echo = TRUE, max.deparse.length = 1000)
```
in the body of the chunk. That method has the disadvantages that bare source code
is included in the knitted file without the accompanying comments, that
specification of the environment through the argument `local` can be
[error-prone](https://pkg.yihui.org/rmarkdown-cookbook/source-script.html)
and that `max.deparse.length` within `source(...)` has to be increased to ensure
that all of the source code is printed (see `help("source")`).

## Knitting documents to HTML and PDF
The following lines can be used to knit documents (i.e., generate output in HTML
or PDF files containing the script with its output and additional text)
containing a file name including a time-stamp. Using this way of creating
documents creates and keeps local R objects in the current environment.
Although that can be useful for inspection and debugging, the potential use of
local objects can lead to non-reproducibility issues, such that the environment
should be cleared before using it and final files should be created by knitting
(e.g., using the `Knit` button in `RStudio`), **not** by using
`rmarkdown::render()` (see
[here](https://pkg.yihui.org/rmarkdown-book/compile.html#fnref2)).

```{r render, eval = FALSE}
DateTimeStamp <- format(Sys.time(), format = "%Y_%m_%d_%H_%M")
rmarkdown::render("FilenameRMarkdownFile.Rmd", output_format = "html_document",
                  output_file = paste0("Filename", DateTimeStamp, ".html"),
                  output_dir = "./knitteddocs")
rmarkdown::render("FilenameRMarkdownFile.Rmd", output_format = "pdf_document",
                  output_file = paste0("Filename", DateTimeStamp, ".pdf"),
                  output_dir = "./knitteddocs")
```

## Adding and collecting code chuncks
Add a new chunk by either (1) clicking the `Insert Chunk` button on the toolbar,
(2) pressing `Ctrl+Alt+I`, or (3) typing the delimiters ```` ```{r} ```` above
and ```` ``` ```` below the code chunk.

To collect all chunks in a R Markdown document that contain R code in a
conventional R script use 
`knitr::purl("FilenameRMarkdownFile.Rmd", documentation = 0)`.

Note that the various R Markdown options are not incorporated in such an
R script, which might hamper exactly reproducing the analyses as executed when
running the R Markdown file, see the note in `help("purl", package = "knitr")`.


## Formatting in R Markdown
### Italic and bold text

````
Using *italic words* and _also italic words_ or **bold words** or even
_An **italic and bold phrase** in italic text_
````
is rendered as:

Using *italic words* and _also italic words_ or **bold words** or even
_An **italic and bold phrase** in italic text_

### Code chunks
R code chunks can be shown with or without its output through argument `eval`
(which is `TRUE` by default):

`````{r code-chunk-format, eval = FALSE}
```{r code-chunk-example}
sum(c(1, 2))
mean(c(1, 2))
```
`````
shows a code chunk with its output, like:
```{r code-chunk-example}
sum(c(1, 2))
mean(c(1, 2))
```

Whereas
`````{r code-chunk-format-no-output, eval = FALSE}
```{r code-chunk-example-no-output, eval = FALSE}
sum(c(1, 2))
mean(c(1, 2))
```
`````
shows a code chunk without its output, like:
```{r code-chunk-example-no-output, eval = FALSE}
sum(c(1, 2))
mean(c(1, 2))
```

### Table
````
| Name       | Description                      |
| :--------- | :--------------------------------|
| `Name one` | Some description                 |
| `Name two` | Some description                 |
````
gives:

| Name       | Description                      |
| :--------- | :--------------------------------|
| `Name one` | Some description                 |
| `Name two` | Some description                 |

### Lists
Lists need an empty line before the first item to be correctly rendered.
Sub-items are created by indenting them. Numbered lists (ordered lists) are
created by using a number followed by a dot (`1.`). Non-numbered lists
(non-ordered lists) are created by using asterisks (`*`), plus signs (`+`), or
dashes (`-`).

````
non-numbered list:

* item one
* item two
  * sub-item one
  * sub-item two

numbered list:

1. item one
2. item two

````
These lists would render as:

non-numbered list:

* item one
* item two
  * sub-item one
  * sub-item two

numbered list:

1. item one
2. item two


### Escaping
To include `n` delimiters ``(`)`` in the **output**, escape them by enclosing
with `n + 1` delimiters:  
for example, ``` `` `some text` `` ```
produces `` `some text` ``,
and ```` ``` `` `some text` `` ``` ````
produces ``` `` `some text` `` ```. Do **not** start a line of code with a
delimiter if you want to escape other delimiters: that would be interpreted as
starting a code section.

To include code to **create** code chunks (i.e., include code which escapes
chunk headings), use
````
r ''````{r use-sum}
1 + 2
```
````
to show how to create a code chunk like
```{r use-sum}
1 + 2
```
